---
title: "Study 1 — The Six Villages (vignette task, spring 2019)"
subtitle: "E1 analysis — habilitace-1, stream #2 (n = 315)"
author: "Jan Nehyba"
date: "2026-07-10"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
---

<!--
  P2 / E1 per analyzy/PLAN-analyz.md. Reproducible source of truth for chapter 5.
  Freezes all cited numbers to ../vystupy/tabulky/ (manifest 05_cisla.csv).
  Run standalone (no quarto):  from analyzy/  ->  Rscript -e 'knitr::purl("notebooks/10_vinety.qmd","/tmp/10_vinety.R"); source("/tmp/10_vinety.R", chdir=TRUE)'
  Data are read-only; this notebook only WRITES to ../vystupy/.
-->

```{r setup}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2)
  library(rstatix); library(effectsize); library(ordinal)
  library(tidyLPA); library(cluster); library(fpc)
})
set.seed(2019)  # collection year; fixed for clusterboot / LPA reproducibility

# --- locate habilitace-1 root robustly (works from analyzy/ or notebooks/) ---
find_root <- function() {
  for (p in c(".", "..", "../..", "../../..")) if (dir.exists(file.path(p, "data/processed"))) return(normalizePath(p))
  stop("Could not locate habilitace-1 root (data/processed not found).")
}
ROOT   <- find_root()
DATA   <- file.path(ROOT, "data", "processed", "vinety_final.csv")
TAB    <- file.path(ROOT, "vystupy", "tabulky")
FIG    <- file.path(ROOT, "vystupy", "obrazky")
dir.create(TAB, showWarnings = FALSE, recursive = TRUE)
dir.create(FIG, showWarnings = FALSE, recursive = TRUE)

# --- principles: variable -> EN label (codebook_vinety.md) ---
V    <- c("v_rovnost","v_nahoda","v_potreby","v_zasluhy","v_rovne_prilezitosti","v_kasty")
LAB  <- c(v_rovnost="Equality", v_nahoda="Chance", v_potreby="Need",
          v_zasluhy="Merit", v_rovne_prilezitosti="Equal opportunity", v_kasty="Caste")

# manifest accumulator: every scalar cited in chapter 5 lands here
MAN <- list()
put <- function(key, val) { MAN[[key]] <<- val; invisible(val) }
```

## Data and integrity check

```{r load}
d <- read.csv(DATA, stringsAsFactors = FALSE)
# Rozšířený vzorek 13. 7. 2026: n=315 (dřív 146; prvních 146 řádků identických, čistá expanze).
N_VIN <- 315
stopifnot(nrow(d) == N_VIN, all(V %in% names(d)))
stopifnot(sum(complete.cases(d[V])) == N_VIN)   # no missing across the six principles
stopifnot(all(as.matrix(d[V]) >= 1 & as.matrix(d[V]) <= 7))

# integrity: means must reproduce the frozen values documented in DATA_README
known <- c(v_rovnost=5.50, v_nahoda=4.50, v_potreby=4.22,
           v_zasluhy=3.93, v_rovne_prilezitosti=2.36, v_kasty=2.00)
obs   <- round(colMeans(d[V]), 2)
stopifnot(all(abs(obs[names(known)] - known) < 0.006))

put("n_total", nrow(d))
put("scale_min", 1); put("scale_max", 7)
cat("n =", nrow(d), "| 0 missing | means match DATA_README\n")
```

## 1. Descriptives and the fairness hierarchy

```{r descriptives}
desc <- d %>%
  select(all_of(V)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "score") %>%
  group_by(var) %>%
  summarise(
    n        = dplyr::n(),
    mean     = mean(score),
    sd       = sd(score),
    median   = median(score),
    iqr      = IQR(score),
    pct_fair   = mean(score >= 5) * 100,   # rated 5-7 (fair side)
    pct_unfair = mean(score <= 3) * 100,   # rated 1-3 (unfair side)
    .groups = "drop"
  ) %>%
  mutate(principle = LAB[var]) %>%
  arrange(desc(mean)) %>%
  select(principle, var, n, mean, sd, median, iqr, pct_fair, pct_unfair)

desc_out <- desc %>% mutate(across(where(is.numeric), ~round(.x, 2)))
write.csv(desc_out, file.path(TAB, "05_vinety_deskriptiva.csv"), row.names = FALSE)
desc_out

# freeze hierarchy endpoints + each principle mean/sd
put("hierarchy_top",    desc$principle[1]);           put("hierarchy_top_mean",    round(desc$mean[1], 2))
put("hierarchy_bottom", desc$principle[nrow(desc)]);  put("hierarchy_bottom_mean", round(desc$mean[nrow(desc)], 2))
for (i in seq_len(nrow(desc))) {
  put(paste0("mean_", desc$var[i]), round(desc$mean[i], 2))
  put(paste0("sd_",   desc$var[i]), round(desc$sd[i], 2))
}
```

## 2. Likert distribution figure

```{r likert-figure}
ord <- desc$var  # highest mean first
likert_df <- d %>%
  select(all_of(V)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "score") %>%
  count(var, score) %>%
  group_by(var) %>% mutate(pct = n / sum(n) * 100) %>% ungroup() %>%
  mutate(principle = factor(LAB[var], levels = rev(LAB[ord])),
         score = factor(score, levels = 1:7))

p_likert <- ggplot(likert_df, aes(principle, pct, fill = score)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "RdYlBu", name = "Rating\n(1 = very unfair,\n7 = very fair)") +
  labs(x = NULL, y = "Percentage of respondents",
       title = "Perceived fairness of six distributive principles",
       subtitle = "Pre-service teachers, spring 2019 (n = 315)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), legend.position = "right")

# Book figure 05_vinety_likert.png is generated ONLY by analyzy/scripts/fig_05_vinety.R
# (mandated theme_book style, CLAUDE.md). This notebook previews the plot inline; it must
# NOT write the book PNG (previously the two generators clobbered each other, last-writer-wins).
p_likert
```

## 3. Omnibus: Friedman test + Kendall's W

```{r friedman}
dl <- d %>%
  mutate(id = factor(row_number())) %>%
  select(id, all_of(V)) %>%
  pivot_longer(-id, names_to = "var", values_to = "score") %>%
  mutate(principle = factor(var, levels = V)) %>%
  arrange(principle, id)   # aligned subject order for paired tests

fr  <- dl %>% friedman_test(score ~ principle | id)
kw  <- dl %>% friedman_effsize(score ~ principle | id)   # Kendall's W

put("friedman_chisq", round(fr$statistic, 2))
put("friedman_df",    fr$df)
put("friedman_p",     signif(fr$p, 3))
put("kendall_w",      round(kw$effsize, 3))
put("kendall_w_mag",  as.character(kw$magnitude))

fr; kw
```

## 4. Post-hoc: pairwise Wilcoxon signed-rank (Holm) + effect size r = Z/sqrt(N)

```{r posthoc}
pw <- dl %>%
  wilcox_test(score ~ principle, paired = TRUE, p.adjust.method = "holm") %>%
  mutate(g1 = LAB[as.character(group1)], g2 = LAB[as.character(group2)])

es <- dl %>%
  wilcox_effsize(score ~ principle, paired = TRUE) %>%
  select(group1, group2, effsize, magnitude)

posthoc <- pw %>%
  left_join(es, by = c("group1", "group2")) %>%
  transmute(pair = paste(g1, "vs", g2),
            n1, n2, statistic,
            p = signif(p, 3), p.adj = signif(p.adj, 3), p.adj.signif,
            # NB: rstatix::wilcox_effsize returns r = Z/sqrt(N), not a rank-biserial correlation;
            # the column name is kept for output compatibility (revision 8 Oct 2026).
            rank_biserial = round(effsize, 3), magnitude)

write.csv(posthoc, file.path(TAB, "05_vinety_posthoc.csv"), row.names = FALSE)
posthoc

put("posthoc_n_pairs",       nrow(posthoc))
put("posthoc_n_sig_holm",    sum(posthoc$p.adj < 0.05))
put("posthoc_n_nonsig_holm", sum(posthoc$p.adj >= 0.05))
```

## 5. Robustness: cumulative link mixed model (ordinal)

```{r clmm}
dclmm <- dl %>% mutate(score_o = ordered(score))
# reference = Equality (top of hierarchy); other principles expected significantly lower
m <- clmm(score_o ~ principle + (1 | id), data = dclmm, Hess = TRUE)
sm <- summary(m)

clmm_fx <- as.data.frame(sm$coefficients)
clmm_fx <- clmm_fx[grepl("^principle", rownames(clmm_fx)), , drop = FALSE]
clmm_fx$principle <- LAB[sub("^principle", "", rownames(clmm_fx))]
clmm_out <- clmm_fx %>%
  transmute(principle,
            estimate = round(Estimate, 3),
            se = round(`Std. Error`, 3),
            z = round(`z value`, 2),
            p = signif(`Pr(>|z|)`, 3))
write.csv(clmm_out, file.path(TAB, "05_vinety_clmm.csv"), row.names = FALSE)
clmm_out

# all contrasts negative (lower than Equality) and their ORDER should match the mean hierarchy
put("clmm_all_below_ref_neg", all(clmm_fx$Estimate < 0))
put("clmm_all_sig",           all(clmm_fx$`Pr(>|z|)` < 0.05))
# does clmm estimate ordering reproduce the descriptive mean ordering (excluding reference)?
mean_ord  <- desc$var[desc$var != "v_rovnost"]
clmm_ord  <- sub("^principle", "", rownames(clmm_fx)[order(clmm_fx$Estimate, decreasing = TRUE)])
put("clmm_reproduces_mean_order", identical(mean_ord, clmm_ord))
```

## 6. Response profiles: exploratory clustering and its stability

```{r clusters}
# Revision 9 Oct 2026: Ward clustering of 7-point ratings has many tied distances, so its
# result depends on the order of rows. The analysis therefore runs in a fixed order
# (sorted by resp_id) and its sensitivity to row order is checked on 200 random permutations.
# (The earlier frozen two-cluster solution, n = 218/97, could not be reproduced from these
# data and this code in any of the permutations; it is superseded by the outputs below.)
d <- d[order(d$resp_id), ]
X <- scale(d[V])                       # z-scored -> cluster on profile shape
dmat <- dist(X)
hc <- hclust(dmat, method = "ward.D2")

# choose k in 2:4 by average silhouette (Ward + PAM)
ks <- 2:4
sil_ward <- sapply(ks, function(k) mean(silhouette(cutree(hc, k), dmat)[, 3]))
sil_pam  <- sapply(ks, function(k) pam(X, k)$silinfo$avg.width)
sil_tab  <- data.frame(k = ks, sil_ward = round(sil_ward, 3), sil_pam = round(sil_pam, 3))
sil_tab

K <- ks[which.max(sil_ward)]
put("cluster_k", K)
put("cluster_sil_ward", round(max(sil_ward), 3))
for (i in seq_along(ks)) put(paste0("cluster_sil_ward_k", ks[i]), round(sil_ward[i], 3))
put("cluster_sil_pam_max", round(max(sil_pam), 3))

# sensitivity to row order: 200 random permutations of the same data
set.seed(2019)
perm <- t(replicate(200, {
  o <- sample(nrow(X)); Xo <- X[o, ]; dmo <- dist(Xo); hco <- hclust(dmo, "ward.D2")
  s <- sapply(ks, function(k) mean(silhouette(cutree(hco, k), dmo)[, 3]))
  cl2 <- cutree(hco, 2)
  dif <- abs(colMeans(Xo[cl2 == 1, , drop = FALSE]) - colMeans(Xo[cl2 == 2, , drop = FALSE]))
  c(k = ks[which.max(s)], top = unname(which.max(dif)))
}))
put("order_perm_n", nrow(perm))
for (k in ks) put(paste0("order_k", k, "_count"), sum(perm[, "k"] == k))
top_tab <- table(factor(V[perm[, "top"]], levels = V))
for (v in V) put(paste0("order_top_sep_", v), as.integer(top_tab[v]))
top_tab

# bootstrap stability (Jaccard) for the solution chosen in the fixed order
cb <- clusterboot(X, B = 1000, clustermethod = hclustCBI,
                  method = "ward.D2", k = K, seed = 2019, count = FALSE)
put("cluster_jaccard_min",  round(min(cb$bootmean), 3))
put("cluster_jaccard_mean", round(mean(cb$bootmean), 3))

# cluster profiles on the ORIGINAL 1-7 scale (fixed order); descriptive only
cl <- cutree(hc, K)
prof <- d %>% mutate(cluster = cl) %>%
  group_by(cluster) %>%
  summarise(across(all_of(V), ~round(mean(.x), 2)), n = dplyr::n(), .groups = "drop")
names(prof) <- c("cluster", LAB[V], "n")
write.csv(prof, file.path(TAB, "05_vinety_cluster_profiles.csv"), row.names = FALSE)
prof
```

## 6b. One hump or two?

```{r continuum}
# Pearson bimodality coefficient: > .555 hints at two modes; <= that => one hump
bimod <- function(x) { x <- x[is.finite(x)]; n <- length(x); m <- mean(x)
  m2 <- mean((x - m)^2); g <- mean((x - m)^3) / m2^1.5; k <- mean((x - m)^4) / m2^2 - 3
  (g^2 + 1) / (k + 3 * (n - 1)^2 / ((n - 2) * (n - 3))) }
put("need_bimodality", round(bimod(d$v_potreby), 2))
put("axis_bimodality", round(bimod(prcomp(X)$x[, 1]), 2))    # first principal component

memb <- data.frame(resp_id = d$resp_id, cluster = cl)
write.csv(memb, file.path(TAB, "05_vinety_cluster_membership.csv"), row.names = FALSE)
cat("bimodality need/axis:", MAN$need_bimodality, "/", MAN$axis_bimodality, "\n")
```

## 7. Exploratory triangulation: latent profile analysis

```{r lpa}
# EXPLORATORY only: LPA triangulates the clustering above. Revision 9 Oct 2026: besides the
# number of classes, the two-class partition itself is compared with the Ward partition.
lpa_ok <- TRUE
lpa_fit <- tryCatch({
  fits <- d %>% select(all_of(V)) %>%
    estimate_profiles(1:4, models = 1)   # model 1: equal variances, no covariances
  get_fit(fits)
}, error = function(e) { lpa_ok <<- FALSE; message("LPA skipped: ", conditionMessage(e)); NULL })

if (lpa_ok && !is.null(lpa_fit)) {
  lpa_tab <- as.data.frame(lpa_fit)[, intersect(c("Classes","BIC","AIC","Entropy"), names(as.data.frame(lpa_fit)))]
  lpa_tab <- lpa_tab %>% mutate(across(where(is.numeric), ~round(.x, 2)))
  write.csv(lpa_tab, file.path(TAB, "05_vinety_lpa_fit.csv"), row.names = FALSE)
  best_lpa <- lpa_tab$Classes[which.min(lpa_tab$BIC)]
  put("lpa_best_classes_bic", best_lpa)
  put("lpa_agrees_with_cluster_k", best_lpa == K)
  print(lpa_tab)

  # the two-class partition: sizes, separating variable, agreement with Ward k = 2
  fit2 <- d %>% select(all_of(V)) %>% estimate_profiles(2, models = 1)
  cl_lpa <- get_data(fit2)$Class
  nl <- table(cl_lpa)
  small <- as.integer(names(which.min(nl)))
  put("lpa2_n_small", as.integer(min(nl)))
  put("lpa2_n_large", as.integer(max(nl)))
  difl <- abs(colMeans(X[cl_lpa == 1, , drop = FALSE]) - colMeans(X[cl_lpa == 2, , drop = FALSE]))
  put("lpa2_top_sep", names(which.max(difl)))
  put("lpa2_small_mean_rovne_prilezitosti", round(mean(d$v_rovne_prilezitosti[cl_lpa == small]), 2))
  put("lpa2_large_mean_rovne_prilezitosti", round(mean(d$v_rovne_prilezitosti[cl_lpa != small]), 2))
  put("lpa2_ward2_ari", round(mclust::adjustedRandIndex(cl_lpa, cutree(hc, 2)), 2))
} else {
  put("lpa_best_classes_bic", NA)
}
```

## 8. Demographics (descriptive only)

```{r demographics}
# demographics available only for a subset; no inferential use (convenience, incomplete)
has_sex <- sum(!is.na(d$pohlavi) & d$pohlavi != "")
has_age <- sum(!is.na(d$vek) & d$vek != "")
put("demo_n_with_sex", has_sex)
put("demo_n_with_age", has_age)
cat("Respondents with sex recorded:", has_sex, "/ 315\n")
cat("Respondents with age recorded:", has_age, "/ 315\n")
```

## 9. Freeze manifest

```{r manifest}
manifest <- data.frame(metric = names(MAN),
                       value  = vapply(MAN, function(x) as.character(x), character(1)),
                       row.names = NULL)
write.csv(manifest, file.path(TAB, "05_cisla.csv"), row.names = FALSE)
cat("Wrote", nrow(manifest), "frozen numbers to 05_cisla.csv\n")
manifest
```
